Kernel Development Using Rust: A Systematic Literature Review

### 2024-05-23 paper.tex
Update on Overleaf.

\documentclass[sigconf]{acmart}
\usepackage{listings, listings-rust}
\usepackage{xcolor}
\usepackage[raggedrightboxes]{ragged2e}
\pagestyle{plain}

@@

The migration of legacy C codebases to memory-safe programming languages like Rust is of paramount
importance in modern software development, aiming to mitigate common security vulnerabilities and
memory-related errors. This paper presents a systematic literature review (SLR) focusing on
strategies and methodologies for integrating Rust into one of the most fundamental areas that are
typically dominated by unsafe languages the operating system kernel. Our study aims to provide a
comprehensive overview of existing research, identify gaps, and suggest future directions in this
domain. Through a rigorous search process, we synthesized relevant studies and extracted key
findings to offer insights into effective approaches for ensuring memory safety when working closley
with hardware.

@@

Once we had our initial collection of papers from the databases we merged all duplicate
papers into one record and then started our first pass which consisted of reading the title and
abstract and either marking the paper as \textit{include} or \textit{exclude}.
Once  the initial screening was completed we were left with 32 potential papers that
we needed an in depth reading.

@@

In this section we present our findings. We address each research question from RQ1 to RQ3.

@@

Chen X et al.~\cite{Chen2023-wb} primary focused on developing a micro-kernel in Rust that could be
formally verified by using both the liner type system or rust and a SMT solver. The new micro-kernel
named Atmosphere follows the tradition of other micro-kernels by pushing most kernel functionally to
users-space thus limiting the surface area that needs to be proved. The authors were able to get a
7.5:1 proof-to-code ratio which is higher that other formerly verified microkernels SeL4 and
CeriKOS, which both have proof-tocode ratio of 19:1 and 20:1 respectively.

@@

\begin{table*}
    \begin{tabular}{||p{2cm}|p{2cm}|p{8cm}||}
    \hline
    Approach & Papers & Operating System\\
    \hline\hline
    Monolithic  & ~\cite{The_kernel_development_community_undated-iw} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
    Micro-kernel & ~\cite{Chen2023-wb} & Atmosphere, Redox, Redleaf\\
    Embedded & ~\cite{Culic2022-bk} & \href{https://github.com/tock/tock}{Tock} \\
    Unikernel & ~\cite{Lankes2019-cm} & RustyHermit \\
    Prototype & ~\cite{Boos2020-zh, Ijaz2023-da} & Thesus \\
    \hline
  \end{tabular}
  \caption{Approaches and Methodologies for Rust in the Kernel}
    \label{tab:RQ1}
\end{table*}

This is a list of operating systems.
\begin{itemize}
\item 
\item 
\item \href{https://hubris.oxide.computer/}{Hubris OS}
\item Theseus (Mono)

\item Splinter

\item Drone OS
\item Bern RTOS

\end{itemize}

@@

@@

Li et al.~\cite{Li2019-ru} explored the feasibility of using Rust in kernel space. The authors took
an existing component, the Out of Memory (OOM) and implemented a replacement using the Rust
programming language. The non-encapsulated interface Rust component which was almost identical to
the original C component only introduced a 0.7\% overhead. The encapsulated Rust component on the
other hand added a 3\% performance overhead. Another important aspect of the kernel is size, we must
be careful to keep the size of the Rust component as close as we can to the C component so we can
still run on all the same hardware. The authors found that there was only a 0.06\% size increase
when compared to the original C implementation.

@@

Ayers et al.~\cite{Ayers2022-sf} focused on the size of rust binaries in the embedded
environment. Their work primary focused on the size of binaries produced using the embedded OS
Tock. The authors identified four major causes of binary growth:

\begin{itemize}
    \item Deeply ingrained monomorphization
    \item Suboptimal compiler generated support code
    \item Hidden data structures and data
    \item Fewer compiler optimizations
\end{itemize}

In addition to the identified causes of binary growth the Ayers et al.~\cite{Ayers2022-sf} have the
following 5 recommendations when using Rust in a size constrained environment:
\begin{itemize}
  \item Minimize Length + Instantiations of Generic Code
  \item Use Trait Objects Sparingly
  \item Don't Panic
  \item Carefully Use Compiler Generated Support Code
  \item Don't use static mut
\end{itemize}

Burtsev et al.~\cite{Burtsev2021-mh} explores what is missing in the Rust programming language to
help solve the isolation problem in an operating system. Currently Rust lacks the ability to express
isolation in the heap without external support. The RedLeaf experiential operating system relied on
an complex interface definition language (IDL) to enforce isolated heaps. The paper enumerates
several properties of the Rust Language that could help with isolation with regards to operating
system development. 1) Supporting trait bounds on functions and closures with any number of
arguments 2) Expose type information in procedural macros 3) Support a collision-free, unique type
identifier 4) Support extendable, no\_std unwind library 5) Support typed assembly language for Rust
6) Support trusted build environments 7) Provide software-only stack guard with extensible probing
interface 8) Develop zero-copy serialization of “plain- old” data structures. The authors argue that
with these changes or inclusions to the Rust programming language developing operating system
kernels would be much easier and safer.

Klimt et al.~\cite{Klimt2023-ob} details the lessons learned in implementing Theseus, an operating
system written in Rust. The authors describe how Theseus uses intralingual design to maximize the
compilers role in enforcing correctness. By leveraging Rusts type system and borrow checker memory
safety and correct ownership transfer can be achieved at a higher level than what could be done in
C. The authors also detailed some of the limitations of intralingual design such as not being as
expressive as many other formal verification techniques due to the limited invariant that can be
enforced by the type system. One of the most important lessons learned was the insight that a linear
type system itself cannot guarantee uniqueness of the resource represented, such when a memory
resources may overlap. The authors introduce the idea of using a hybrid approach of verification
where they leverage both the linear type system and an SMT solver.

Klimt et al.~\cite{Klimt2023-ob} explore the challenges of using rust to write an operating
system. It is impossible to write a complete operating system in 100\% safe rust. For example when
writing a memory management system raw pointers must be used in order to modify the hardware. The
authors found four typical soundness issues in their code base. 1) Unsynchronized Global State - any
use of mutable statics is unsafe, 2) C-Style Abstractions - Use rust style abstractions to properly
encapsulate interal unsafe usage of raw pointers, 3) Aliasing Mutable References - Giving out raw
pointers to memory that is also referenced mutably, and 4) Reimplementing Memory Access - instead of
accessing specific memory regions in assembly, creating and using references to the whole region is
preferable. The authors also explore the bootstrapping problem with rust systems. How do we ensure
that ownership of memory is sound in the operating system itself? In operating systems written in C
the kernel provides the ownership root to applications running on top, finding a new ownership root
is a open research question with regards to using Rust for operating system development.

>>>>

\documentclass[sigconf]{acmart}
\usepackage{listings, listings-rust}
\usepackage{xcolor}
\usepackage[raggedrightboxes]{ragged2e}

\pagestyle{plain}

@@

The migration of legacy C code bases to memory-safe programming languages like Rust is of paramount importance in modern software development, aiming to mitigate common security vulnerabilities and memory-related errors. This paper presents a systematic literature review (SLR) focusing on strategies and methodologies for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe languages the operating system kernel. Our study aims to provide a comprehensive overview of existing research, identify gaps, and suggest future directions in this domain. Through a rigorous search process, we synthesized relevant studies and extracted key findings to offer insights into effective approaches for ensuring memory safety when working closely with hardware.

@@

Once we had our initial collection of papers from the databases we merged all duplicate papers into one record and then started our first pass which consisted of reading the title and abstract and either marking the paper as \textit{include} or \textit{exclude}. Once  the initial screening was completed we were left with 29 potential papers that we needed an in depth reading.

@@

In this section we present our findings. We address each research question from \textbf{RQ1} - \textbf{RQ3}.

@@

Chen X et al.~\cite{Chen2023-wb} primary focused on developing a micro-kernel in Rust that could be formally verified by using both the liner type system or rust and a SMT solver. The new micro-kernel named Atmosphere follows the tradition of other micro-kernels by pushing most kernel functionally to users-space thus limiting the surface area that needs to be proved. The authors were able to get a 7.5:1 proof-to-code ratio which is higher that other formerly verified microkernels SeL4 and CeriKOS, which both have proof to code ratio of 19:1 and 20:1 respectively.

@@

Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel. The authors proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel what can run programs written in any language that can compile down to WASM.

Liang et al.~\cite{Liang2021-bo} developed a new operating system named Rustpi with a micro-kernel architecture implemented in Rust to explore how modern language features can help with building reilabe systems.

\begin{table}
    \begin{tabular}{||l|l|l||}
    \hline
    Approach & Papers & Operating System in Rust\\
    \hline\hline
    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe} & Atmosphere, Redox, Redleaf\\
    Embedded & \cite{Culic2022-bk} & \href{https://github.com/tock/tock}{Tock}, \href{https://hubris.oxide.computer/}{Hubris}, \href{https://www.drone-os.com/}{Drone}, \href{https://bern-rtos.org/}{Bern} \\
    Unikernel & \cite{Lankes2019-cm},  \cite{Boos2020-zh}, \cite{Ijaz2023-da}  & RustyHermit, Theseus \\
    Exokernel & \cite{Li2024-yb} & W-Kernel \\
    \hline
  \end{tabular}
  \caption{Approaches and Methodologies for Rust in the Kernel}
    \label{tab:RQ1}
\end{table}

@@

Comparing performance, throughput, latency, and resource utilization between different systems is very difficult. For example several of the operating systems written in Rust such as Tock are a complete rethinking of how an operating system is designed and Rust for Linux (RFL) is still relatively new and has not had the benefit of decades of incremental performance improvements like components that are written in C. Additionally, direct comparisons between a Monolithic kernel, Micro-kernel, and Unikernel, or comparisons between a Real Time Operation System (RTOS) and a General Purpose operating System are not directly meaningful due to the vastly different goals of each system. Therefore, we will focus more on the overall efforts that are specific to the Rust programming language and the challenges presented. We summarize our findings in table~\ref{tab:RQ2}.

@@

Li et al.~\cite{Li2024-be} explored the feasibility of using Rust in kernel space. The authors took
an existing component, the Out of Memory (OOM) and implemented a replacement using the Rust
programming language. The non-encapsulated interface Rust component which was almost identical to
the original C component only introduced a 0.7\% overhead. The encapsulated Rust component on the
other hand added a 3\% performance overhead. Another important aspect of the kernel is size, we must
be careful to keep the size of the Rust component as close as we can to the C component so we can
still run on all the same hardware. The authors found that there was only a 0.06\% size increase
when compared to the original C implementation.

\begin{table}
    \begin{tabular}{||l|l|l||}
    \hline
    No. & Implication & Studies that Reported the challenge\\
    \hline\hline
    1 & Performance & \cite{Gonzalez2023-ek}, \cite{Li2024-be}, \cite{Ma2023-ef}\\
    2 & Throughput & \cite{Gonzalez2023-ek}\\
    3 & Latency & \cite{Culic2022-bk} \\
    4 & Resource Utilization & 0 \\
    \hline
  \end{tabular}
  \caption{Performance Implications of using Rust in the Kernel}
    \label{tab:RQ2}
\end{table}

@@

\begin{table*}[htbp]
    \begin{tabular}{p{1cm} p{5cm} p{10cm}}
        \hline
        No. & Challenge & Description\\
        \hline
        1 & Binary Size~\cite{Ayers2022-sf}  & 
        \begin{itemize}
            \item Deeply ingrained monomorphization which increased the size of the rust binaries
            \item Compiler optimizations are not as mature as some C based compilers thus increasing binary size
            \item Hidden data structures and data
            \item Sub-optimal compiler generated support code
        \end{itemize}
        \\
        \hline
        2 & Missing Features~\cite{Burtsev2021-mh} &
        \begin{itemize}
            \item Supporting trait bounds on functions and closures with any number of arguments
            \item Expose type information in procedural macros
            \item Support a collision-free, unique type identifier

            \item Support typed assembly language for Rust
            \item Support trusted build environments 
            \item Provide software-only stack guard with extensible probing interface
            \item Develop zero-copy serialization of “plain- old” data structures
        \end{itemize}
        \\
        \hline
        3 & Soundness~\cite{Klimt2023-ob} &
        \begin{itemize}
            \item Unsynchronized Global State - any use of mutable statics is unsafe
            \item C-Style Abstractions - Use rust style abstractions to properly encapsulate internal unsafe usage of raw pointers
            \item Aliasing Mutable References - Giving out raw pointers to memory that is also referenced mutably
            \item Re-implementing Memory Access - instead of accessing specific memory regions in assembly, creating and using references to the whole region is preferable
        \end{itemize}
        \\
        \hline
        4 & Panics \cite{Ma2023-ef},\cite{Burtsev2021-mh} &
        \begin{itemize}
            \item Support extendable, no\_std unwind library
            \item Stack unwinding in embedded environments
        \end{itemize}
        \\
    \hline
  \end{tabular}
  \caption{Challenges Unique to the Rust Programming Language}
    \label{tab:RQ3}
\end{table*}

Ayers et al.~\cite{Ayers2022-sf} focused on the size of rust binaries in the embedded environment. Their work primary focused on the size of binaries produced using the embedded OS Tock. In addition to the identified causes of binary growth the authors have the following 5 recommendations when using Rust in a size constrained environment:
\begin{itemize}
  \item Minimize Length + Instantiations of Generic Code
  \item Use Trait Objects Sparingly
  \item Don't Panic
  \item Carefully Use Compiler Generated Support Code
  \item Don't use static mut
\end{itemize}

Burtsev et al.~\cite{Burtsev2021-mh} explores what is missing in the Rust programming language to help solve the isolation problem in an operating system. Currently Rust lacks the ability to express isolation in the heap without external support. The RedLeaf experiential operating system relied on an complex interface definition language (IDL) to enforce isolated heaps. The paper enumerates several properties of the Rust Language that could help with isolation with regards to operating system development that are detailed in table~\ref{tab:RQ3}. The authors argue that with these changes or inclusions to the Rust programming language developing operating system kernels would be much easier and safer.

Klimt et al.~\cite{Klimt2023-ob} details the lessons learned and challenges when implementing Theseus, an operating system written in Rust. First they found that it is impossible to write a complete operating system in 100\% safe rust. For example when writing a memory management system raw pointers must be used in order to modify the hardware. Despite the limitations imposed by the hardware the authors describe how Theseus leveraged intralingual design to maximize the compilers role in enforcing correctness. By leveraging Rusts type system and borrow checker memory safety and correct ownership transfer can be achieved at a higher level than what could be done in C. The authors also detailed some of the limitations of intralingual design such as not being as expressive as many other formal verification techniques due to the limited invariant that can be enforced by the type system. One of the most important lessons learned was the insight that a linear type system itself cannot guarantee uniqueness of the resource represented, such when a memory resources may overlap. The authors introduce the idea of using a hybrid approach of verification where they leverage both the linear type system and an SMT solver. The authors also explore the bootstrapping problem with rust systems. How do we ensure that ownership of memory is sound in the operating system itself? In operating systems written in C the kernel provides the ownership root to applications running on top, finding a new ownership root is a open research question with regards to using Rust for operating system development.
